---
title: "ISUH 2022 Physical Activity Workshop"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# call packages

We first load all the required packages

```{r warning = F}
# Load tidyverse (install if not loaded)
if (!require("tidyverse")) {
  install.packages("tidyverse")
  library(tidyverse)
}

# Install DRPA package (for calculating relative risks)
if (!require("drpa",character.only = TRUE)) {
  print('Installing "drpa" package...')
  remotes::install_github("meta-analyses/drpa")
  library(drpa)
  print("")
}

```

# Constant values

Marginal Metabolic Equivalent Task (mMET) values associated with walking
and cycling can be obtained from the compendium of physical activities
(<https://sites.google.com/site/compendiumofphysicalactivities/home>).
The values defined below are from ?? and ?? respectively.

Please replace the "PATH_TO..." strings with the full path to the
mortality and travel survey data, downloadable from ??. Alternatively,
leave these as defined and use setwd() to change your working directory
to the folder containing the data.

```{r}
WALK_MMET = 2.5
CYCLE_MMET = 5.6
PATH_TO_MORTALITY_DATA = "gbd_sao_paulo.csv"
PATH_TO_TRAVEL_SURVEY = "travel_pa_data.csv"
```

# Import travel survey

The travel survey dataset for Sao Paulo is a single-day travel diary in
which each row represents a trip, or trip stage, made by the respondent.
Public transport trips are broken down into stages to identify each mode
that is used (for example, walking to your home bus stop, taking one
bus, and then walking from the destination stop to work is recorded as a
3-stage trip). This stage data is useful for identifying walking and
cycling physical activity that takes place as a part of public transport
trips.

The travel survey dataset has been linked with a physical activity
dataset containing the individual's total non-transport physical
activity. This information in contained in the variable "ltpa_marg_met".

[Individual-level variables]{.underline}

-   **participant_id:** Unique ID for each individual

-   **age:** participant's age as an integer

-   **sex:** male or female

-   **ltpa_marg_met:** leisure-time (non-transport) physical activity in
    marginal metabolic equivalent task hours (mMET-hrs) per week

[Trip-level variables:]{.underline}

-   **trip_id:** Unique ID for each trip made by each individual

-   **trip_mode:** Main mode of the trip

-   **trip_distance:** Total trip distance in kilometers

[Stage-level variables:]{.underline}

-   **stage_mode:** Mode of the trip stage

-   **stage_duration:** Duration of each stage in minutes

```{r message=F}
# Load travel survey data
travel_survey <- read_csv(PATH_TO_TRAVEL_SURVEY) %>% arrange(participant_id,trip_id)

# View first few rows
head(travel_survey)
```

# Import mortality data

Mortality data for Sao Paulo has been obtained from the global burden of
disease database (<https://ghdx.healthdata.org/gbd-2019>). For this
exercise we will only consider mortality from all causes. Therefore, we
filter the dataset to only deaths from all causes. Mortality can be
calculated by dividing the number deaths by the total population in each
group. Furthermore, we change sex to all lower case to match the values
in the travel survey.

```{r}
# Read mortality data
mortality_data <- read_csv(PATH_TO_MORTALITY_DATA)

# View first few rows
head(mortality_data)

# Filter down to all-cause mortality and update/remove variable names
mortality_data <- mortality_data %>%
  filter(cause_name == "All causes", 
         measure_name.x == "Deaths") %>%
  mutate(sex = tolower(sex_name),
         age_gr = age_name,
         deaths = val,
         mortality = deaths / population) %>% 
  select(-cause_name,-measure_name.x,-sex_name,-age_name,-val)
```

# Aggregate to the individual and calculate total marginal met hours per week

In this step we use dplyr to aggregate trip and stage level data to
determine total walking and cycling physical activity for each
individual. Some variables are already at the individual level (e.g.
age, sex, ltpa_marg_met), so we keep these variables in the aggregated
dataset using the "first" function. We combine age into groups to allow
linking with mortality data.

Note that as the travel diary is only for a single day, so it must be
multiplied by 7 to create an estimate of physical activity throughout
the week. Leisure-time physical activity (represented using the variable
"ltpa_marg_met") is already a weekly estimate so this does not need to
be multiplied by 7.

```{r}
individual_data <- travel_survey %>%
  group_by(participant_id) %>%
  summarise(age = first(age),
            sex = first(sex),
            leisure_mmet_hr = first(ltpa_marg_met),
            walk_minutes = 7 * sum(stage_duration[stage_mode == "pedestrian" | stage_mode == "walk_to_pt"]),
            bike_minutes = 7 * sum(stage_duration[stage_mode == "bike"])) %>%
  mutate(age_gr = case_when(age < 5 ~ "0 to 4",
                            age < 10 ~ "5 to 9",
                            age < 15 ~ "10 to 14",
                            age < 20 ~ "15 to 19",
                            age < 25 ~ "20 to 24",
                            age < 30 ~ "25 to 29",
                            age < 35 ~ "30 to 34",
                            age < 40 ~ "35 to 39",
                            age < 45 ~ "40 to 44",
                            age < 50 ~ "45 to 49",
                            age < 55 ~ "50 to 54",
                            age < 60 ~ "55 to 59",
                            age < 65 ~ "60 to 64",
                            age < 70 ~ "65 to 69"),
         walk_mmet_hr = WALK_MMET * walk_minutes / 60,
         cycle_mmet_hr = CYCLE_MMET * bike_minutes / 60)

# View first few rows
head(individual_data)
```

# Calculate total marginal metabolic equivalent task hours and define scenarios

Here, we calculate the total weekly physical activity (in mMET-hrs per
week) by summing physical activity from walking, cycling, and
(non-transport) leisure time.

We define a scenario in which transport physical activity is reduced by
half, while (non-transport) leisure time physical activity remains the
same.

```{r}
# Calculate baseline and scenario physical activity
individual_data <- individual_data %>% 
  mutate(total_mmet_hr_BASE = leisure_mmet_hr + walk_mmet_hr + cycle_mmet_hr,
         total_mmet_hr_SCEN = leisure_mmet_hr + walk_mmet_hr/2 + cycle_mmet_hr/2)

# Show summary statistics
individual_data %>% select(total_mmet_hr_BASE, total_mmet_hr_SCEN) %>% summary()

# Plot physical activity distributions using empirical density curves
plot_data <- individual_data %>% 
  pivot_longer(cols = c(total_mmet_hr_BASE, total_mmet_hr_SCEN), 
               names_to = "scenario", names_prefix = "total_mmet_hr_", values_to = "mmet_hrs")

ggplot(plot_data, aes(x = mmet_hrs, colour = scenario)) + geom_density() + scale_x_log10() + 
   xlab("mMET-hours per week") + ylab("cumulative share of individuals")
```

# Calculate relative risks

We can use the "dose_response" function from the DRPA package to
estimate the relative risk associated with each individual's dose of
physical activity, given in units of marginal metabolic equivalent tasks
per week.

Alternatively the dose-response curve data can be download from
<https://shiny.mrc-epid.cam.ac.uk/meta-analyses-physical-activity/> and
sampled manually.

```{r}
## Calculate relative risk for baseline and scenario
individual_data <- individual_data %>%
    mutate(rr_base = dose_response(cause = "all-cause-mortality", 
                                   outcome_type = "fatal", 
                                   dose = total_mmet_hr_BASE)[,1],
           rr_scen = dose_response(cause = "all-cause-mortality", 
                                   outcome_type = "fatal", 
                                   dose = total_mmet_hr_SCEN)[,1])

# Plot dose-response curves using an emperical cumulative density function (ECDF)
plot_data <- individual_data %>% 
  pivot_longer(cols = c(rr_base, rr_scen), 
               names_to = "scenario", names_prefix = "rr_", values_to = "value")

ggplot(plot_data, aes(x = value, colour = scenario)) + stat_ecdf() + 
   xlab("All-cause mortality relative risk") + ylab("cumulative share of individuals")
```

As visible in the plot, the maximum RR is 1.0 (associated with zero
physical activity) and the minimum is 0.6755 (associated with PA volumes
of 12.84 mMET-hours per week or higher). Relative risks for the scenario
are higher than those for the baseline due to the reduction in physical
activity in the scenario.

# Mortality

Use mortality rates from the GBD to investigate changes in total
mortality in the survey population.

```{r}
# Link travel survey data to gbd mortality data
individual_data <- individual_data %>% 
  left_join(select(mortality_data,age_gr,sex,mortality)) %>% 
  mutate(mortality_base = mortality * rr_base,
         mortality_scen = mortality * rr_scen)


SURVEY_POPULATION_MORTALITY_BASE = sum(individual_data$mortality_base)
SURVEY_POPULATION_MORTALITY_BASE

SURVEY_POPULATION_MORTALITY_SCEN = sum(individual_data$mortality_scen)
SURVEY_POPULATION_MORTALITY_SCEN

SURVEY_POPULATION_MORTALITY_DIFFERENCE = SURVEY_POPULATION_MORTALITY_SCEN - SURVEY_POPULATION_MORTALITY_BASE
SURVEY_POPULATION_MORTALITY_DIFFERENCE
```

These calculations show that in the survey population of 10.000, the
scenario results in 1,19 more deaths than baseline.

# PIF

In this step we use the survey data to calculate population impact
factors (PIFs) for the whole population as well as for various subgroups

```{r}
# Define function to calculate PIFs
calculate_PIF <- function(data) {
  baseline <- sum(data[["rr_base"]])
  scenario <- sum(data[["rr_scen"]])
  return((baseline - scenario)/baseline)
}

# Calculate PIFs for the whole population
PIF = individual_data %>% calculate_PIF()

# Calculate PIFs for different population subgroups
PIF_female <- individual_data %>% filter(sex == "female") %>% calculate_PIF()
PIF_male <- individual_data %>% filter(sex=="Male") %>% calculate_PIF()

PIF_under50 <- individual_data %>% filter(age < 50) %>% calculate_PIF()
PIF_over50 <- individual_data %>% filter(age >= 50) %>% calculate_PIF()

# Calculate PIFs for ALL population subgroups from the GBD
population_groups <- individual_data %>%
  group_by(age_gr,sex) %>%
  summarise(count = n(),
            PIF = (sum(rr_base) - sum(rr_scen))/sum(rr_base)) %>%
  left_join(select(mortality_data,age_gr,sex,deaths)) %>%
  mutate(change = PIF * deaths)

# View fist few rows
head(population_groups)

# Calculate total change in mortality
TOTAL_MORTALITY_CHANGE = sum(population_groups$change)
TOTAL_MORTALITY_CHANGE
```
