---
title: "ISUH 2022 Physical Activity Workshop"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

Physical Activity workshop for ISUH 2022 Conference in Valencia, 24
October 2022

Jardín Botánico (Aula Cavanilles), 16:30-18:00

# SETUP: load packages

We first load all the required packages

```{r warning = F}
# Load tidyverse (install if not loaded)
if (!require("tidyverse")) {
  install.packages("tidyverse")
  library(tidyverse)
}

# Install DRPA package (for calculating relative risks)
if (!require("drpa",character.only = TRUE)) {
  print('Installing "drpa" package...')
  remotes::install_github("meta-analyses/drpa")
  library(drpa)
  print("")
}

```

# Exercise 1: Working with the drpa package

Here are some usage example for the dvra package. The dose-response
curves in this package are non-linear spline curves created from a
meta-analysis of longitudinal cohort studies considering all
non-occupational physical activity.

In addition to this R package, the dose-response curves can also be
visualized and download from
<https://shiny.mrc-epid.cam.ac.uk/meta-analyses-physical-activity/>.

Exercise questions:

1.  How do the shapes of the curves for CVD (fatality vs non-fatality)
    differ? *(hint: use the link to the shiny app)*

2.  For depression, how do 'combined fatality and non-fatality' relative
    risks differ for 5 and 10 mMET/wk (hovering the mouse over the
    curve) *(hint: type "?drpa::dose_response" for help on using this
    function)*

3.  For neurological diseases, use downloaded data from the shiny app
    how to investigate do mortality relative risks differ for 5 and 10
    mMET/wk? *(hint: use the link to the shiny app and use your own
    application to open)*

4.  Using the r package function, determine cancer mortality RR for
    people accumulating 5, 12 and 18 mMETs/week

```{r}

# Some examples
drpa::dose_response(cause = "all-cause-mortality", outcome_type = "fatal", dose = 5)

drpa::dose_response(cause = "depression", outcome_type = "fatal-and-non-fatal", dose = c(5,12,18))

```

# Exercise 2: Microsimulation Health Impact Assessment

## Constant values

Marginal Metabolic Equivalent Task (mMET) values associated with walking
and cycling can be obtained from the compendium of physical activities
(<https://sites.google.com/site/compendiumofphysicalactivities/home>).
For this exercise we use 2.5 mMET for walking and 5.6 mMET for cycling.

```{r}
WALK_MMET = 2.5
CYCLE_MMET = 5.6
MORTALITY_DATA_FILE = "gbd_sao_paulo.csv"
TRAVEL_SURVEY_DATA_FILE = "travel_pa_data.csv"
```

## Import travel survey

The travel survey dataset for Sao Paulo is a single-day travel diary in
which each row represents a trip, or trip stage, made by the respondent.
Public transport trips are broken down into stages to identify each mode
that is used (for example, walking to your home bus stop, taking one
bus, and then walking from the destination stop to work is recorded as a
3-stage trip). This stage data is useful for identifying walking and
cycling physical activity that takes place as a part of public transport
trips.

The travel survey dataset has been linked with a physical activity
dataset containing the individual's total non-transport physical
activity. This information in contained in the variable "ltpa_marg_met".

[Individual-level variables]{.underline}

-   **participant_id:** Unique ID for each individual

-   **age:** participant's age as an integer

-   **sex:** male or female

-   **ltpa_marg_met:** leisure-time (non-transport) physical activity in
    marginal metabolic equivalent task hours (mMET-hrs) per week

[Trip-level variables:]{.underline}

-   **trip_id:** Unique ID for each trip made by each individual

-   **trip_mode:** Main mode of the trip

-   **trip_distance:** Total trip distance in kilometers

[Stage-level variables:]{.underline}

-   **stage_mode:** Mode of the trip stage

-   **stage_duration:** Duration of each stage in minutes

```{r message=F}
# Load travel survey data
travel_survey <- read_csv(TRAVEL_SURVEY_DATA_FILE) %>% arrange(participant_id,trip_id)

# View first few rows
head(travel_survey)
```

## Import mortality data

Mortality data for Sao Paulo has been obtained from the global burden of
disease database (<https://ghdx.healthdata.org/gbd-2019>). For this
exercise we will only consider mortality from all causes. Therefore, we
filter the dataset to only deaths from all causes. Mortality can be
calculated by dividing the number deaths by the total population in each
group. Furthermore, we change sex to all lower case to match the values
in the travel survey.

```{r}
# Read mortality data
mortality_data <- read_csv(MORTALITY_DATA_FILE)

# View first few rows
head(mortality_data)

# Filter down to all-cause mortality and update/remove variable names
mortality_data <- mortality_data %>%
  filter(cause_name == "All causes", 
         measure_name.x == "Deaths") %>%
  mutate(sex = tolower(sex_name),
         age_gr = age_name,
         deaths = val,
         mortality = deaths / population) %>% 
  select(-cause_name,-measure_name.x,-sex_name,-age_name,-val)
```

## Aggregate trip physical activity to the individual

In this step we use dplyr to aggregate trip and stage level data to
determine total walking and cycling physical activity for each
individual. Some variables are already at the individual level (e.g.
age, sex, ltpa_marg_met), so we keep these variables in the aggregated
dataset using the "first" function. We combine age into groups to allow
linking with mortality data.

Note that as the travel diary only covers a single day, so it must be
multiplied by 7 to create an estimate of physical activity throughout
the week. Leisure-time physical activity (represented using the variable
"ltpa_marg_met") is already a weekly estimate so this does not need to
be multiplied by 7.

```{r}
individual_data <- travel_survey %>%
  group_by(participant_id) %>%
  summarise(age = first(age),
            sex = first(sex),
            leisure_mmet_hr = first(ltpa_marg_met),
            walk_minutes = 7 * sum(stage_duration[stage_mode == "pedestrian" | stage_mode == "walk_to_pt"]),
            bike_minutes = 7 * sum(stage_duration[stage_mode == "bike"])) %>%
  mutate(age_gr = case_when(age < 5 ~ "0 to 4",
                            age < 10 ~ "5 to 9",
                            age < 15 ~ "10 to 14",
                            age < 20 ~ "15 to 19",
                            age < 25 ~ "20 to 24",
                            age < 30 ~ "25 to 29",
                            age < 35 ~ "30 to 34",
                            age < 40 ~ "35 to 39",
                            age < 45 ~ "40 to 44",
                            age < 50 ~ "45 to 49",
                            age < 55 ~ "50 to 54",
                            age < 60 ~ "55 to 59",
                            age < 65 ~ "60 to 64",
                            age < 70 ~ "65 to 69"),
         walk_mmet_hr = WALK_MMET * walk_minutes / 60,
         cycle_mmet_hr = CYCLE_MMET * bike_minutes / 60,
         transport_mmet_hr = walk_mmet_hr + cycle_mmet_hr)

# View first few rows
head(individual_data)
```

## Define scenarios and calculate total mMET hours

Here, we calculate the total weekly physical activity (in mMET-hrs per
week) by summing physical activity from walking, cycling, and
(non-transport) leisure time.

We define a scenario in which transport physical activity is reduced by
half, while (non-transport) leisure time physical activity remains the
same.

```{r}
# Define Scenario
individual_data <- individual_data %>%
  mutate(transport_mmet_hr_BASE = transport_mmet_hr,
         transport_mmet_hr_SCEN = transport_mmet_hr / 2)


# Calculate baseline and scenario physical activity
individual_data <- individual_data %>% 
  mutate(total_mmet_hr_BASE = leisure_mmet_hr + transport_mmet_hr_BASE,
         total_mmet_hr_SCEN = leisure_mmet_hr + transport_mmet_hr_SCEN)

# Show summary statistics
individual_data %>% select(total_mmet_hr_BASE, total_mmet_hr_SCEN) %>% summary()

# Plot physical activity distributions using empirical density curves
plot_data <- individual_data %>% 
  pivot_longer(cols = c(total_mmet_hr_BASE, total_mmet_hr_SCEN), 
               names_to = "scenario", names_prefix = "total_mmet_hr_", values_to = "mmet_hrs")

ggplot(plot_data, aes(x = mmet_hrs, colour = scenario)) + geom_density() + scale_x_log10() + 
   xlab("mMET-hours per week") + ylab("cumulative share of individuals")
```

## Calculate relative risks

In this step, we estimate the relative risk associated with each
individual's volume of physical activity at baseline and the scenario.
We consider two methods:

**Method 1:** We use the "dose_response" function from the DRPA package
as described previously. Recall that these dose-response curves are
non-linear and consider non-occupational physical activity, **including
transport and leisure.**

**Method 2:** We use linear dose-response curves based on the guidance
in the WHO Health Economic Assessment Tool for walking and cycling. The
guidance considers transport physical activity independently, and
therefore this method **does not consider leisure-time activities.**

The plots below visualizes the distribution of relative risks across the
survey population, comparing baseline to the scenario and both RR
calculation methods.

```{r}
## METHOD 1: Calculate RR using non-linear dose-response function
individual_data <- individual_data %>%
    mutate(rr_base = dose_response(cause = "all-cause-mortality", 
                                   outcome_type = "fatal", 
                                   dose = total_mmet_hr_BASE)[,1],
           rr_scen = dose_response(cause = "all-cause-mortality", 
                                   outcome_type = "fatal", 
                                   dose = total_mmet_hr_SCEN)[,1])

## METHOD 2: Calculate RR using a linear function based on the WHO HEAT tool
heat_dose_response <- function(mMET) {
  return(pmax(0.6755,1 - ((1 - 0.9)/11.25) * mMET))
}

individual_data <- individual_data %>%
  mutate(rr_base_linear = heat_dose_response(transport_mmet_hr_BASE),
         rr_scen_linear = heat_dose_response(transport_mmet_hr_SCEN))

## Plot dose-response curves using an emperical cumulative density function (ECDF)
plot_data <- individual_data %>% 
  pivot_longer(cols = c(rr_base, rr_scen, rr_base_linear, rr_scen_linear), 
               names_to = "scenario", names_prefix = "rr_", values_to = "value") %>%
  mutate(type = factor(case_when(endsWith(scenario,"linear") ~ "linear",
                                 TRUE ~ "nonlinear"),levels = c("nonlinear","linear")),
         scenario = case_when(scenario == "base_linear" ~ "base",
                              scenario == "scen_linear" ~ "scen",
                              TRUE ~ scenario))

ggplot(plot_data, aes(x = value, colour = scenario, linetype = type)) + stat_ecdf() + 
   xlab("All-cause mortality relative risk") + ylab("cumulative share of individuals")
```

As visible in the plot, the maximum RR is 1.0 (associated with zero
physical activity) and the minimum is 0.6755 (associated with PA volumes
of 12.84 mMET-hours per week or higher). Relative risks for the scenario
are higher than those for the baseline due to the reduction in physical
activity in the scenario.

## Estimate mortality using GBD data

Use mortality rates from the GBD to investigate changes in total
mortality in the survey population.

```{r}
# Link travel survey data to gbd mortality data
individual_data <- individual_data %>% 
  left_join(select(mortality_data,age_gr,sex,mortality), by = c("sex","age_gr")) %>% 
  mutate(mortality_base = mortality * rr_base,
         mortality_scen = mortality * rr_scen)


SURVEY_POPULATION_MORTALITY_BASE = sum(individual_data$mortality_base)
SURVEY_POPULATION_MORTALITY_SCEN = sum(individual_data$mortality_scen)
SURVEY_POPULATION_MORTALITY_DIFFERENCE = SURVEY_POPULATION_MORTALITY_SCEN - SURVEY_POPULATION_MORTALITY_BASE

print(paste("Baseline mortality in survey population =",SURVEY_POPULATION_MORTALITY_BASE))
print(paste("Scenario mortality in survey population =",SURVEY_POPULATION_MORTALITY_SCEN))
print(paste("Change in mortality for survey population =",SURVEY_POPULATION_MORTALITY_DIFFERENCE))

```

These calculations show that in the survey population of 10.000, the
scenario results in 1,19 more deaths than baseline.

## Calculate Population Impact Factors (PIFs)

In this step we use the survey data to calculate population impact
factors (PIFs) for the whole population as well as for various subgroups

```{r}
# Define function to calculate PIFs
calculate_PIF <- function(data) {
  baseline <- sum(data[["rr_base"]])
  scenario <- sum(data[["rr_scen"]])
  return((baseline - scenario)/baseline)
}

# Calculate PIFs for the whole population
PIF = individual_data %>% calculate_PIF()

# Examples: PIFs for different population subgroups
PIF_female <- individual_data %>% filter(sex == "female") %>% calculate_PIF()
PIF_male <- individual_data %>% filter(sex=="Male") %>% calculate_PIF()

PIF_under50 <- individual_data %>% filter(age < 50) %>% calculate_PIF()
PIF_over50 <- individual_data %>% filter(age >= 50) %>% calculate_PIF()

# Calculate PIFs for ALL population subgroups from the GBD
population_groups <- individual_data %>%
  group_by(age_gr,sex) %>%
  summarise(count = n(),
            PIF = (sum(rr_base) - sum(rr_scen))/sum(rr_base)) %>%
  left_join(select(mortality_data,age_gr,sex,deaths)) %>%
  mutate(change = PIF * deaths)

# View fist few rows
head(population_groups)

# Calculate total change in mortality
TOTAL_MORTALITY_CHANGE = sum(population_groups$change)
print(paste("Change in mortality for total population =",TOTAL_MORTALITY_CHANGE))

```

For the full population of the Brazilian state of Sao Paulo, this
modelling exercise predicts that halving current physical activity
levels will lead to about 4830 additional deaths per year (of a total
population of \~ 45 million).
